---
title: Some further zoning analyses for Madison
author: Harald Kliems
date: '2019-06-25'
slug: some-further-zoning-analyses-for-madison
categories: []
tags:
  - R
  - zoning
  - landuse
  - Madison
---
I got some good feedback after posting my article on Madison's zoning restriction on Facebook and Reddit, prompting some additional analyses.

# Why do you look only at residentially zoned land?
Several people pointed out that comparing the zoning districts that only allow detached single-family homes (SFR) to all *residential* zoning districts may be misleading. In many cities the densest parts are not in residential zoning districts but in downtown areas zoned for mixed use. I think that's a fair point, but it was the original NYT article that chose only residential areas as the comparator, and I wanted to make Madison comparable to the other cities in the NYT analysis. 

Nonetheless, I quickly ran the analysis comparing SFR to any district except "special districts." These includes parks, conservancies, agricultural land, airports, or the UW-Campus---in other words districts that are highly unlikely to ever have residential use. 
```{r eval = FALSE}
other <- c("A",
           "UA",
           "CN",
           "PR",
           "AP",
           "CI",
           "PD",
           "PHMP")

#detached single-family only
res_sfr <- data %>%
  filter(ZONING_CODE %in% sfr_zones) %>%
  summarize(sum(ShapeSTArea))

# only special district area
other_area <- data %>%
  filter(ZONING_CODE %in% other) %>%
  summarise(sum(ShapeSTArea))

total <- data %>% summarize(sum(ShapeSTArea))
res_sfr/(total-other_area)
```

The result? 49% of all land that reasonably could have residential units on it allows only single-family attached housing. Since there is no point of comparison to other cities, it's hard to interpret that number.

# Is there a correlation between the percentage of SFR zoning and overall density?
It's a good question: Does the metric the NYT used actually have an association with overall density of a city? (Note that even if there is a correlation, this absolutely doesn't mean there is a causal connection! There are many, many factors that impact a city's overall density.)

There are few data points to work with. I retrieved density information for all the NYT cities and Madison from Wikipedia:

```{r eval = FALSE}
dens_zon <- read_csv("data/density_zoning.csv")
```

Let's see what the overall correlation is:
```{r eval = FALSE}
 dens_zon %>% 
   spread(Measure, Value) %>%
   summarize(cor(density, sfr))
```
Hm, -0.885. That's a very, very high correlation. But it's always good to look at your data points in a scatterplot to see what's actually going on:
 
```{r eval = FALSE}
 dens_zon %>% 
   spread(Measure, Value) %>%
   ggplot(aes(sfr, density, label = City)) +
   geom_point() +
   geom_text(check_overlap = TRUE, angle = 45, vjust = 0.1, hjust = -0.1, size = 3) +
   scale_y_continuous(limits = c(0, 35000)) +
   scale_x_continuous(expand = c(0, 0.2)) +
   xlab("Proportion detached single family zoning to all residential zoning") +
   ylab("Overall density (population/sq mi)")
```



> # let's take out NYC
> dens_zon %>% 
+   spread(Measure, Value) %>%
+   filter(City != "New York") %>%
+   summarize(cor(density, sfr))
# A tibble: 1 x 1
  `cor(density, sfr)`
                <dbl>
1              -0.693
> 
> # let's also take out DC
> dens_zon %>% 
+   spread(Measure, Value) %>%
+   filter(City != "New York" & City != "Washington") %>%
+   summarize(cor(density, sfr))
# A tibble: 1 x 1
  `cor(density, sfr)`
                <dbl>
1              -0.366
> 

and then created a scatterplot.

