---
title: What Does Density Look Like in Madison?
author: Harald Kliems
date: '2020-01-30'
slug: what-does-density-look-like-in-madison
categories: []
tags:
  - landuse
  - Madison
  - R
draft: true  
---
This is a follow-up to my previous post on *where* Madison is most dense. But what do these dense areas look like? Inspired by the ["Every Lot" bot]( http://fakeisthenewreal.org/everylot/), I'll use the Google Street View API via the `googleway` package to (hopefully) create some visual sense of density. 
```{r}
## Code from previous post to get 10 most dense block groups in Dane County
library(sf)
library(tidycensus)
library(tidyverse)
library(ggplot2)
library(leaflet)
library(googleway)
key <- Sys.getenv("STREET_VIEW_API_KEY") #key requires a credit card on file

pop <- get_acs(geography = "block group", 
                     variables = "B01003_001", 
                     state = "WI",
                     county = "Dane",
                     geometry = TRUE,
                     keep_geo_vars = TRUE)
pop2 <- pop %>% 
  mutate(pop_density = estimate / (ALAND /27878400)) %>%  #create variable for population density
  drop_na %>%
  filter(pop_density > 100000) %>% 
  arrange(desc(pop_density))                     
```

A good starting point for exploration would be the [centroid](https://en.wikipedia.org/wiki/Centroid#Of_a_polygon) of the block group. Block groups can have odd shapes, though, which can lead to the true centroid ending up outside the polygon. I'll use `st_point_on_surface` instead.

```{r}
pop2 <- pop2 %>% 
  mutate(center = st_point_on_surface(geometry))
```
A quick visual check:
```{r}
library(ggmap)
ggplot() +
  geom_sf(data = pop2,
          aes(fill = pop_density), 
          alpha = 0.5,
          color = NA) +
  geom_sf(data = pop2,
          aes(geometry = center))
```
Looks good! Now let's get Street View images. Testing with the first row:

```{r}
google_streetview(location = st_coordinates(pop2$center[1]),
    size = c(400,400), output = "plot",
    key = key)
```
"No imagery here" -- that's because lat/long are in the reverse order! 
```{r}
centers <- st_coordinates(pop2$center)
centers2 <- centers[, rev(seq_len(ncol(centers)))]
google_streetview(location = centers2[1,],
    size = c(400,400), output = "plot",
    key = key)
```

